Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-26 Manuscript/manuscript.Rmd
added to discussion

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

@@

In summary, we considered combinations of data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on chromosome-specific genomic bins (excluding chromosome 9), *C = {CHR1, ..., CHR22}*, using 5-fold cross-validation. Model inputs were evaluated by assessing the balanced accuracy (BA) of each combination. <!--The results presented are for the ARROWHEAD boundaries, with PEAKACHU boundaries presented in the Supplementary material. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |S| \times |C|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 \times 21 (= 3360)$ models.--> [??? When we have all results, Peakachu results will likely need to be described in parallel. They are about loops, theoretically different from TAD boundaries. For discussion]

![**A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the chromosome-specific rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)). Step 2 is to split **A** into a training and testing set according to a 7:3 ratio. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to implement the random forest algorithm and perform 5-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png)

@@

![**Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal predictive performance for GM12878.** Line plots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.](Figures/figure3_performances.png)

@@

<!--Additional performance metrics were reported for both cell lines including positive predictive value (PPV; precision), F1-Score (F1), Accuracy, Area Under the Receiver Operating Characteristic Curve (AUC), Area Under the Precision-Recall Curve.-->

Here we present *preciseTAD*, a random forest-based approach coupled with density-based spatial clustering (DBSCAN) and scalable partition around medoids (PAM/CLARA) to more precisely predict TAD or loop boundaries on the linear genome. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a user defined TAD(loop) calling tool and ChIP-seq defined genomic annotations as the feature space. We optimized our RF model by creating a computational framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedure, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors. The method was benchmarked against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from the GM12878 and K562 cell lines.

@@

@@

<!-- Discuss advantages and disadvantages of Balanced Accuracy. The disadvantage is that we restrict TP to the originally detected TAD boundaries. While there are likely many more that couldn't be detected but are TP. We currently count them as FP -->

@@

>>>>

Our framework is resolution-specific and uses called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

@@

In summary, we considered combinations of data from a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ chromosomes, while reserving the $i^{th}$ chromosome for testing. Five-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model inputs were evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (See Methods). <!--The results presented are for the ARROWHEAD boundaries, with PEAKACHU boundaries presented in the Supplementary material. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |S| \times |C|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 \times 21 (= 3360)$ models.--> [??? When we have all results, Peakachu results will likely need to be described in parallel. They are about loops, theoretically different from TAD boundaries. For discussion]

![**A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the chromosome-specific rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ chromosomes. Step 2 is to use the binned data matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 5-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png)

@@

![**???[needs updating with new framework, takes longest, do last]Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal predictive performance for GM12878.** Line plots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.](Figures/figure3_performances.png)

@@

Here we present *preciseTAD*, a random forest-based approach coupled with spatial clustering (DBSCAN) and scalable partition around medoids (CLARA) that allows for more precise prediction of TAD or loop boundaries. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a user defined TAD(loop) calling tool and ChIP-seq defined genomic annotations as the feature space. We optimized our RF model by creating a computational framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedure, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors. We benchmarked *preciseTAD* against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we also applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from the GM12878 and K562 cell lines.

@@

Besides balanced accuracy (BA), we investigated five other performance metrics to evaluate the optimal combination of data level characteristics for boundary region prediction including accuracy, area under the receiver operating characteristic curve (AUC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 6-7). The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benifits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling.

@@

@@

**Supplementary Table 6.**

**Supplementary Table 7.**
